---
title: "Measuring cumulative disease burden with the AUC method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{msAUC Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

```{r setup}
library(msAUC)
set.seed(2024)
```

## Why an area-under-the-curve estimand?

In trials with a progressive, ordinal outcome -- for example successive
thresholds of kidney-function decline, end-stage renal disease, and death -- a
conventional *time-to-first-event* analysis collapses each participant to a
single endpoint and ignores everything that happens afterwards. Two patients who
both reach a first event at the same time are treated identically even if one
stabilises while the other progresses to dialysis and dies.

The **area-under-the-curve (AUC)** estimand instead tracks the *mean cumulative
severity score* over the whole follow-up window. Each participant accrues score
as they pass through increasingly severe states, and the area under that mean
curve, restricted to `[0, tau]`, summarises the total burden experienced. The
treatment effect is reported as a **ratio** and a **difference** of these areas.

This vignette walks through a complete analysis on simulated data.

## Data format

`msAUC` expects data in long multistate format, the same layout used by the
**rmt** package. Each subject contributes one or more rows; `status` is coded

- `0` for censoring,
- `1, ..., K` for the ordinal non-fatal states in increasing severity,
- `K + 1` for death.

A subject's records end with either a death row (`status == K + 1`) or a
censoring row (`status == 0`).

`sim_cumburden()` generates data in exactly this form. It uses a continuous-time
progressive multistate model: each subject starts event-free and, from whatever
state they currently occupy, faces two competing exponential clocks -- one to
progress to the next more-severe state, one to die. Whichever fires first
happens, and the process repeats until death, random dropout, or administrative
censoring at `tau`.

The treatment effect is induced through the `hr` argument, which multiplies the
**transition intensities** (the instantaneous rates of progressing and of dying)
in the active arm. It is therefore a hazard ratio defined per transition, not
per composite endpoint: `hr < 1` slows every transition, so treated patients
linger longer in milder states and accrue less cumulative burden. (If you need
the treatment to act differently on progression and on mortality, set
`hr_progression` and `hr_death` separately; if you want the progression and
death rates to differ by disease state -- for example, faster decline and higher
mortality as patients worsen -- pass vectors to `progression` and `mortality`;
and the
allocation ratio is set with `ratio` (1:1 by default) -- see `?sim_cumburden`.)

Here we simulate a trial with four non-fatal states plus death, applying a
common hazard ratio of 0.78 to every transition in the active arm:

```{r simulate}
dat <- sim_cumburden(n = 800, K = 4, tau = 6, hr = 0.78, seed = 1)
head(dat, 8)

# status 0 = censored, 1-4 = ordinal non-fatal states, 5 = death
table(dat$status)
```

Each subject has a single treatment assignment that is constant across their
rows:

```{r arms}
arm_by_id <- tapply(dat$arm, dat$id, function(x) x[1])
table(arm = arm_by_id)
```

## Fitting the model

`auc_fit()` estimates the AUC in each arm and contrasts them. The only required
argument beyond the data is the restriction time `tau`:

```{r fit}
fit <- auc_fit(dat, tau = 6)
fit
```

The printout reports:

- the **AUC** in each arm (cumulative time lost to progression),
- the **AUC ratio** (active / control); below 1 means lower burden under
  treatment,
- the **AUC difference** (active − control), in score-years; below 0 means
  lower burden under treatment.

## Decomposing the effect

`summary()` adds a per-state decomposition: how much event-free time the
treatment adds at each severity level. These contributions sum exactly to the
absolute AUC reduction.

```{r summary}
summary(fit)
```

The `estimate` column is the extra event-free time (in years) the treatment
provides for "reaching state *s* or worse". Reading down the states shows
whether the benefit is concentrated in milder progression or extends all the way
to the most severe outcomes.

You can pull any quantity out of the fitted object directly:

```{r extract}
fit$ratio
fit$difference
fit$decomposition
```

## Visualising the burden

`auc_plot()` reproduces the central figure for this estimand: the expected
cumulative score in each arm (primary axis) with the AUC ratio as a function of
the restriction time overlaid (secondary, dashed). The gap between the two solid
curves *is* the burden the treatment averts.

```{r plot}
auc_plot(fit)
```

If you would rather build the figure yourself, `auc_curve_data()` returns the
underlying curves and ratio as plain data frames:

```{r curve-data}
cd <- auc_curve_data(fit)
str(cd, max.level = 2)
```

## Choosing which endpoints to include

A recurring question is how sensitive the conclusion is to the *definition* of
the composite. The `states` argument lets you include any subset of the ordinal
states; the retained states are automatically re-scored `1, ..., m` in
increasing severity, with death always carrying the top score.

```{r sensitivity}
# All five states (default)
all5 <- auc_fit(dat, tau = 6)

# Drop the mildest state
drop1 <- auc_fit(dat, tau = 6, states = c(2, 3, 4, 5))

# The two most severe states only (e.g. ESRD + death)
severe <- auc_fit(dat, tau = 6, states = c(4, 5))

sens <- data.frame(
  definition = c("all 5 states", "drop mildest", "ESRD + death"),
  n_states   = c(all5$n_states, drop1$n_states, severe$n_states),
  ratio      = c(all5$ratio["estimate"], drop1$ratio["estimate"], severe$ratio["estimate"]),
  ci_low     = c(all5$ratio["conf.low"], drop1$ratio["conf.low"], severe$ratio["conf.low"]),
  ci_high    = c(all5$ratio["conf.high"], drop1$ratio["conf.high"], severe$ratio["conf.high"])
)
sens
```

Comparing the AUC ratio across endpoint definitions is a transparent
sensitivity analysis: a conclusion that holds whether or not the mildest states
are included is more robust than one that depends on them.

## Choosing the restriction time

`tau` should reflect the clinically relevant follow-up horizon and must lie
within the observed event times. Because the AUC accumulates over `[0, tau]`,
larger `tau` generally yields a larger absolute difference; reporting the ratio
alongside the difference helps separate the magnitude of burden from the
relative effect.

```{r tau-sweep}
taus <- c(3, 4, 5, 6)
do.call(rbind, lapply(taus, function(tt) {
  f <- auc_fit(dat, tau = tt)
  data.frame(tau = tt,
             ratio = round(f$ratio["estimate"], 3),
             difference = round(f$difference["estimate"], 3))
}))
```

## Notes and limitations

- **Scoring.** The included states are scored with equal one-unit steps (1 for
  the mildest state you keep, up to `m` for death). This R package does not have
  a function to customize severity weights.
- **State skipping and interval censoring.** A subject first observed in a
  severe state -- or who dies with no milder event recorded -- is treated as
  reaching every lower threshold at that same time. Because disease states are
  typically assessed only at scheduled visits, the recorded time to a milder
  threshold is really the time it was *first detected* (at that level or
  worse), which can be later than when it was actually crossed.
- **Two arms.** `auc_fit()` compares exactly two arms and treats them as
  independent when combining variances.

## References

- Claggett B, Tian L, Fu H, et al. Quantifying the totality of treatment effect
  with multiple event-time observations in the presence of a terminal event.
  *Statistics in Medicine* 2018; 37(25): 3589--3598.
- Sun J, et al. Improve the precision of area under the curve
  estimation for recurrent events through covariate adjustment. *Statistics in
  Medicine* 2025; 44(15--17): e70187.
